---
title: "pm2.5 eda"
output:
  pdf_document: default
  html_document: default
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(data.table)
library(dplyr)
library(dtplyr)
library(stringr)
```

EDA done based on questions for Course Project 2 for Exploratory Data Analysis course by John Hopkins University

## Loading data

```{r loading}
dt <- fread("data/merged.csv") # run prep_data to create this file
```

Data consists of a summary data frame and a codetable data frame, which have been merged into merged.csv. Summary contains all of the PM2.5 commissions data for 1999, 2002, 2005, 2008. The code table provides a mapping from the SCC digit strings in the summary table to the actual name of the PM2.5 source.

```{r}
summary(dt)
```

```{r}
str(dt)
```

```{r}
slice_sample(dt, n=5)
```

## Question 1: Have total emissions from PM2.5 decreased from 1999 to 2008?

```{r, echo=FALSE}
with(dt, {
    sums <- tapply(Emissions, year, sum)
    years <- as.numeric(names(sums))
    plot(sums ~ years,
         xaxt="n",
         main="Total emission trend",
         xlab="Year",
         ylab="Total pm2.5 emissions (tons)")
    
    # fit <- lm(sums ~ years)
    abline(lm(sums ~ years))
    
    axis(1, at=unique(year), labels=unique(year))
})
```

This plot shows a clear negative trend, meaning that emissions have decreased from 1999 to 2008.

## Question 2: Have total emissions from PM2.5 decreased in Baltimore City, Maryland

```{r, echo=FALSE}
bmore <- subset(dt, dt$fips == "24510")
with(bmore, {
    sums <- tapply(Emissions, year, sum)
    years <- as.numeric(names(sums))
    plot(sums ~ years,
         xaxt="n",
         main="Total emission trend, Baltimore City",
         xlab="Year",
         ylab="Total pm2.5 emissions (tons)")
    
    fit <- lm(sums ~ years)
    abline(fit)
    
    axis(1, at=unique(year), labels=unique(year))
})

```

By subsetting data to Baltimore City, Maryland (fips == "24510") we can we that total emissions have raised here.

## Question 3: Of the 4 types of sources, which see decreases in emissions for Baltimore City

```{r echo=FALSE, message=FALSE, warning=FALSE}
processed <- bmore %>% group_by(type, year) %>% summarise(emissions_sum = sum(Emissions))

ggplot(processed, aes(year, emissions_sum)) +
    facet_grid(.~type) +
    geom_point() +
    geom_smooth(method="lm", se=FALSE) +
    ggtitle("Trends per type ", subtitle="Baltimore City") +
    xlab("Year") +
    ylab("Total pm2.5 emissions (tons)") +
    scale_x_continuous(labels=unique(processed$year), breaks=unique(processed$year)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

From this plot we can see taht the pm2.5 trends downwards for NON-ROAD, NON-POINT, and ROAD. But it is trending up for the POINT type.

## Question 4: How have emissions from coal combustion sources changed?

```{r,echo=FALSE, message=FALSE, warning=FALSE}
coal <- dt[str_detect(EI.Sector, "[cC]oal")] %>% group_by(year) %>% summarise(emissions_sum = sum(Emissions))

ggplot(coal, aes(year, emissions_sum)) +
    geom_point() +
    geom_smooth(method="lm", se=FALSE, aes(colour="1999-2008")) +
    geom_smooth(method="lm", se=FALSE, data=head(coal, 3), aes(colour="1999-2005")) +
    ggtitle("Coal emissions trend") +
    xlab("Year") +
    ylab("Total pm2.5 emissions (tons)") +
    scale_x_continuous(labels=unique(processed$year), breaks=unique(processed$year)) +
    guides(colour=guide_legend(title="Trend over"))

```

Yes! Although the trend for 1999-2005 (shown in red) seems to be fairly consistent, the only real decrease being 2008.

## Question 5: How have emissions from motor vehicle sources changed?

```{r echo=FALSE, message=FALSE, warning=FALSE}
motor <- bmore[str_detect(EI.Sector, "Mobile")] %>% group_by(year, EI.Sector) %>% summarise(emissions_sum = sum(Emissions))
total <- motor %>% group_by(year) %>% summarise(emissions_sum = sum(emissions_sum)) %>% mutate(EI.Sector = "TOTAL")
all <- rbind(total, motor)

ggplot(all, aes(year, emissions_sum, colour=EI.Sector)) +
    geom_point() +
    geom_smooth() +
    ggtitle("Vehicle emissions trend", "Baltimore City") +
    xlab("Year") +
    ylab("Total pm2.5 emissions (tons)") +
    scale_x_continuous(labels=unique(processed$year), breaks=unique(processed$year))

```

Overall emissions from vehicles are lower than in 1999, but have risen in 2008 mainly due to commercial marine vessels, which is the main contributor.

## Question 6: Has Baltimore City emissions from motor vehicles changed greater than in Los Angeles County?

```{r echo=FALSE, message=FALSE, warning=FALSE}
filtered <- dt[fips == "24510" | fips == "06037"]
sums <- filtered[str_detect(EI.Sector, "Mobile")] %>%
    group_by(year, fips) %>%
    summarise(emissions_sum = sum(Emissions)) %>%
    mutate(county=case_match(fips,
                             "24510" ~ "Baltimore City",
                             "06037" ~ "Los Angeles County"))

ggplot(sums, aes(year, emissions_sum, colour=county)) +
    geom_point() +
    geom_smooth() +
    ggtitle("Vehicle emissions trend", "Baltimore City vs Los Angeles County") +
    xlab("Year") +
    ylab("Total pm2.5 emissions (tons)") +
    scale_x_continuous(labels=unique(processed$year), breaks=unique(processed$year))

```

The change in Los Angeles County is much greater, presumably due to the significant population difference (9.78 million vs 560 thousand as of 2008).
